Mergeable Types for Highly-Available Replicated Data

### 2017-02-23 paper/semantics.tex
LCA Semantics

When merging two concurrent versions $v_1$ and $v_2$, the common
ancestor argument for \C{merge} must the LCA of $v_1$ and $v_2$,
without which \C{merge} may yeild unexpected results. This is
demonstrated for the grow-only counter in
Fig.~\ref{fig:merge-needs-lca}, where incorrect count is obtained if a
common ancestor that is not an LCA is used to merge 4 and 7. While in
this example there is a unique LCA for 4 and 7, in general this may
not be the case. With unrestrained branching and merging, there is no
bound on the number of LCAs a pair of versions can have.
For example, in
Fig.~\ref{fig:criss-cross-lcas}, the merge of 1 with 3 is preceded by
two ``criss-cross'' merges between their respective branches resulting
in there being two LCAs (3 and 4) for 1 and 3. Multiple LCAs can
occur even without criss-cross merges, as demonstrated by
Fig.~\ref{fig:external-lcas}.

>>>>

When merging two concurrent versions $v_1$ and $v_2$, the common
ancestor argument for \C{merge} must the LCA of $v_1$ and $v_2$,
without which \C{merge} may yeild unexpected results. This is
demonstrated for the grow-only counter in
Fig.~\ref{fig:merge-needs-lca}, where incorrect count is obtained if a
common ancestor that is not an LCA is used to merge 4 and 7. While in
this example there is a unique LCA for 4 and 7, in general this may
not be the case. With unrestrained branching and merging, there is no
bound on the number of LCAs a pair of versions can have.
For example, in
Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded by
two ``criss-cross'' merges between their respective branches resulting
in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs can occur
even without criss-cross merges, as demonstrated by
Fig.~\ref{fig:external-lcas}. The problem with concurrent versions
with multiple LCAs is that they do not lend themselves to three-way
merging. This problem also arises in the context of source
control systems, and often solved via ad-hoc mechanisms.
GitHub~\cite{github}, for instance, recursively merges LCAs to compute
a virtual ancestor, which then serves as the LCA for merging the
concurrent versions. This method is demonstrated for the example in
Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
merged (with LCA of 10) to generate -1 as the virtual LCA to merge 5
and 4. A major downside with this method is that it makes no
guarantees of the relationship between the virtual ancestor and the
concurrent versions; the former may not even be a legal ancestor of
the latter as per the semantics of the data type. For instance, the
integer type in Fig.~\ref{fig:criss-cross-lcas} may in fact represent
a bank account, which disallows any activity on the account if balance
is ever known to be less than zero. Thus, from the perspective of a
bank account, it doesn't make sense how concurrent versions 3 and 0
emerged from -1, since the only transition allowed by the semantics
from -1 is to itself. Clearly, ad hoc mechanisms, which work for the
text, may not work for more sophisticated data types.

Fortunately, unlike the source control systems where branching
structure is entirely dictated by the user, \name abstracts away the
branching structure from the programmer, hence retains the ability to
constrain it in a way that it deems fit. In particular, \name solves
the problem of multiple LCAs by suitably constraining the branching
structure such that the problem never arises. The constraints are
imposed either implicitly, as a result of how operational semantics
defines an atomic step, or explicitly, by insisting that certain
conditions be met before merging a pair of versions
(\rulelabel{E-Pull-Wait}). Firstly, the operational semantics already
disables criss-cross merges since it only ever merges versions that
are latest on their respective branches. For instance let $v_{11}$ and
$v_{21}$ be the latest versions on branches $b_1$ and $b_2$,
respectively. Meging $b_2$ into $b_1$ requires entails $v_{21}$ into
$v_{11}$ to generate version $v_{12}$ on $b_1$. Now, merging $b_1$
into $b_2$ translates to merging $v_{12}$ into $v_{21}$ (a
\emph{fast-forward} merge, in Git parlance), but not $v_{11}$ into
$v_{21}$, thus preventing a criss-cross branching structure. In other
words, a criss-cross branching structure is prevented due to the
order among merges between conflicting branches introduced as a result
of \rulelabel{E-Pull-Wait} being an atomic step.

Secondly, we impose certain pre-conditions on the merging branches to
preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
intuition is as follows: consider the branch $b_1$ at the instance of
merging $b_3$. Since it has already merged $b_2$, a version on $b_2$
($v_{22}$) could be a common ancestor for a version on $b_1$
($v_{12}$) and some other version (call it $v$). Now, if $b_1$ merges
$b_3$, same could be true of $b_3$ and $b_1$: a version on $b_3$
($v_{32}$) could be a common ancestor for the new version on $b_1$
($v_{13}$), and the other version $v$. Since $v_{22}$ is also an
ancestor of $v_{13}$, and both ancestors are not ordered by the
ancestor relation, $v_{13}$ and $v$ end up with two LCAs. In
Fig.~\ref{fig:external-lcas}, the role of $v$ is played by the version
$v_{44}$. We observe that this scenario can be prevented if, when
merging $b_3$, $b_1$ insists on an ancestor relation between the
merged version ($v_{22}$) of the $b_2$, the previously merged branch,
and the latest version of $b_3$, the current merging branch. We call
the last merged version the \emph{external locus} of the branch. By
requiring that, for every branch $b$, the external locii of $b$ at
various points in time (i.e., external locus of every prefix of $b$)
be totally ordered, we effectively enforce the total order among
versions that have more than one common ancestor, thereby preventing
the possibility of multiple LCAs.

%% Multiple versions that merge any pair of versions are totally
%% ordered. That is, if $v_1$ and $v_2$ are ancestors of $v_3$ and
%% $v_4$, then $v_3$ and $v_4$ are ordered by the ancestor relation.
